Exploring the ages of rotating stars using galactic dynamics: a novel approach to calibrating gyrochronology

### 2020-02-25 paper/results.tex
Rocio's comments

The top panel of figure \ref{fig:vplot} shows rotation period versus effective
temperature for the \mct\ sample, coloured by the standard deviation of their
(\vb) velocities, where \sigmavb\ was calculated for groups of stars over a
grid in $\log_{10}$(period) and temperature.
If we assume that mass dependent heating does not strongly affect this sample
and \vb\ at low galactic latitudes is an unbiased tracer of \vz, then \vb\
velocity dispersion can be interpreted as an age proxy, and stars plotted in a
similar color in figure \ref{fig:vplot} are similar ages.
\begin{figure}
  \caption{
    Top: Rotation period vs effective temperature for stars in the \mct\
    sample, colored by the velocity dispersions of stars calculated over a
    grid in $\log_{10}$(period) and \teff.
    Black lines show gyrochrones from a gyrochronology model that projects the
    rotation-color relation of
    Praesepe to longer rotation periods over time \citep{angus2019}.
    These gyrochrones do not appear to reflect the evolution of field stars at
    long rotation periods/old ages: they do not trace lines of constant
    velocity dispersion.
    Gyrochrones are plotted at 0.5, 1, 1.5, 2, 2.5, 4 and 4.57 Gyrs in both top
    and bottom panels.
    Bottom: Same as top panel with rotation period vs {\it mass}
    \citep[from the Kepler Input Catalog][]{brown2011}.
    White lines show gyrochrones from a model that includes mass and
    age-dependent angular momentum transport between the core and envelope
    \citep{spada2019}.
    Qualitatively, these gyrochrones reflect the evolution of field
    stars at long rotation periods/old ages: they trace lines of constant
    velocity dispersion by reproducing periods of `stalled' rotational
    evolution for K-dwarfs.
}
  \centering
    \includegraphics[width=1\textwidth]{main_figure}
\label{fig:vplot}
\end{figure}
Overall, figure \ref{fig:vplot} shows that velocity dispersion increases with
rotation period across all temperatures, implying that rotation period
increases with age as expected.
Black lines show gyrochrones from the \citet{angus2019} gyrochronology model,
which projects the rotation-color relation of Praesepe to longer rotation
periods over time.
These gyrochrones are plotted at 0.5, 1, 1.5, 2, 2.5, 4 and 4.57 Gyrs.
At the youngest ages, these gyrochrones describe the data well: the palest
yellow (youngest) stars with the lowest velocity dispersions all fall close to
the 0.5 Gyr gyrochrone.
% Lines of constant age (isochrones) appear to follow the shape of the
% Praesepe-based gyrochronology model (black dotted line) at young ages.
% However, at older ages it appears that the relation between rotation period
% and \teff\ flattens out, until eventually rotation period {\it decreases} with
% decreasing effective temperature at a given age.
However, although the 0.5 Gyr and 1 Gyr gyrochrones also trace constant
velocity dispersion/age among the field stars, by 1.5 Gyrs the gyrochrones
start to {\it cross} different velocity dispersion regimes.
For example, the 1.5 Gyr gyrochrone lies on top of stars with velocity
dispersions of around 10-11 kms$^{-1}$ at 5000-5500K and stars with $\sim$15
\kms\ velocity dispersions at 4000-4500K.
The gyrochrones older than 1.5 Gyr also cross a range of velocity dispersions.
If these were true isochrones however, they should follow lines of constant
velocity dispersion.
At ages older than around 1 Gyr, it appears that gyrochrones should have a more
flattened, or even inverted relationship between rotation period and effective
temperature than these Praesepe-based models.
% For late K-dwarfs older than around 1 Gyr, it appears that lower-mass stars
% rotate more rapidly than higher-mass stars.
% This is the opposite of the trend that is observed in young open clusters such
% as Praesepe and the Hyades.

The bottom panel of figure \ref{fig:vplot} shows velocity dispersion as a
function of rotation period and {\it mass}, obtained from the Kepler Input
Catalog \citep{brown2011}, with gyrochrones from the \citep{spada2019}
gyrochronology model shown as white lines.
These gyrochrones are also plotted at 0.5, 1, 1.5, 2, 2.5, 4 and 4.57 Gyrs.
Although these models do not trace lines of constant velocity dispersion at
0.5 and 1 Gyr, they do appear to reproduce trends in the data at ages of 1.5
Gyr and older.
The \citet{spada2019} models appear to qualitatively agree with the data:
their rotation period-\teff\ relation flattens out over time and eventually
inverts.

@@

To investigate whether mass-dependent heating could be acting on the \kepler\
sample, we selected late K and M dwarfs observed by both \kepler\ and \gaia,
whose MS lifetimes exceed around 11 Gyrs and are therefore representative of
the initial mass function.
We could not perform this analysis on the \mct\ sample, because it only
includes stars with {\it detectable} rotation periods, and since lower-mass
stars stay active for longer it is likely that it contains a strong mass-age
correlation.
We selected all \kepler\ targets with dereddened \gaia\ \gcolor\ colors
greater than 1.2 (corresponding to an effective temperature $\lesssim$
4800 K) and absolute \gaia\ $G$-band magnitudes $<$ 4.
We also eliminated visual binaries by removing stars above a 6th order
polynomial, fit to the MS on the \gaia\ CMD (similar to the one shown in
figure \ref{fig:age_gradient}).
We then applied the quality cuts described above in section
\ref{sec:the_data}.
To search for evidence of mass-dependent heating we calculated the (\vb)
velocity dispersion of stars in effective temperature bins.
Sigma clipping was performed at 3$\sigma$ to remove high and low velocity
outliers before calculating the standard deviation of stars in each bin.
These extreme velocity outliers may be very old late K and M dwarfs, or they
result from using \vb\ instead of \vz, which introduces additional velocity
scatter.

@@

Having found no strong evidence for mass-dependent heating, we next tested
the validity of \vb\ as a proxy for \vz\ in more detail.
At a galactic latitude, $b$, of zero, $v_b=v_z$, however for increasing values
of $b$, this equivalence becomes an approximation that grows noisier with $b$.
To test the validity of the \vb$\sim$\vz\ approximation over a range of
latitudes we downloaded stellar data from the \Gaia\ Universe Model Snapshot
(GUMS) simulation -- a simulated \Gaia\ catalog \citep{robin2012}.
We downloaded stars from four pointings in the \kepler\ field with galactic
latitudes of around 5\degrees, 10\degrees, 15\degrees, and 20\degrees, out to
a limiting magnitude of 16 dex, and calculated their \vz\ and \vb\ velocities.
The relationship between \vz\ and \vb\ is close to 1:1, with \vz\ greater than
\vb\ by around 4.5 kms$^{-1}$ at $b=5$, due to the Sun's own motion in the
Galaxy.
We subtracted this offset and examined the residuals of the \vz\ -- \vb\
relationship to investigate the variance as a function of Galactic latitude.
We found that \vb\ is drawn from a heavy-tailed distribution, centered on \vz,
with standard deviation increasing with $b$ (see figure \ref{fig:vb_vz}).
The standard deviation of \vz-\vb\ was around 3kms$^{-1}$ at $b \sim 5^\circ
$, 4kms$^{-1}$ at 10$^\circ$, 6kms$^{-1}$ at 15$^\circ$, and 9kms$^{-1}$ at
20$^\circ$.

@@

Because of the noisy relationship between \vb\ and \vz\, in this paper we do
not attempt to convert velocity dispersion (\sigmavb) into an age via an
age-velocity dispersion relation (AVR) \citep[\eg][]{holmberg2009}.
Although we find that \sigmavb\ can be used to rank groups of stars by age, a
more careful analysis that includes formal modeling of the \vb\ -- \vz\
relationship will be needed to calculate absolute ages.
% For example, the velocity distributions could be modeled as a mixture of
% Gaussians in order to account for the additional velocity dispersion caused by
% the \vz-\vb\ transformation.
The RVs of most of these stars will become available in \Gaia\ DR3, allowing
calculations of \vz, which can be used to calculate more reliable ages via an
AVR and suitable modeling of the selection function.

>>>>

The top panel of figure \ref{fig:vplot} shows rotation period versus effective
temperature for the \mct\ sample, coloured by the standard deviation of their
(\vb) velocities, where \sigmavb\ was calculated for groups of stars over a
grid in $\log_{10}$(period) and temperature.
If we assume that mass dependent heating does not strongly affect this sample
and \vb\ at low galactic latitudes is an unbiased tracer of \vz, then \vb\
velocity dispersion can be interpreted as an age proxy, and stars plotted in a
similar color in figure \ref{fig:vplot} are similar ages.
\begin{figure}
  \caption{
    Top: Rotation period vs effective temperature for stars in the \mct\
    sample, colored by the velocity dispersions of stars calculated over a
    grid in $\log_{10}$(period) and \teff.
    Black lines show gyrochrones from a gyrochronology model that projects the
    rotation-color relation of
    Praesepe to longer rotation periods over time \citep{angus2019}.
    These gyrochrones do not appear to reflect the evolution of field stars at
    long rotation periods/old ages: they do not trace lines of constant
    velocity dispersion.
    Gyrochrones are plotted at 0.5, 1, 1.5, 2, 2.5, 4 and 4.57 Gyr in both top
    and bottom panels.
    Bottom: Same as top panel with rotation period vs {\it mass}
    \citep[from the Kepler Input Catalog][]{brown2011}.
    White lines show gyrochrones from a model that includes mass and
    age-dependent angular momentum transport between the core and envelope
    \citep{spada2019}.
    Qualitatively, these gyrochrones reflect the evolution of field
    stars at long rotation periods/old ages: they trace lines of constant
    velocity dispersion by reproducing periods of `stalled' rotational
    evolution for K-dwarfs.
}
  \centering
    \includegraphics[width=1\textwidth]{main_figure}
\label{fig:vplot}
\end{figure}
Overall, figure \ref{fig:vplot} shows that velocity dispersion increases with
rotation period across all temperatures, implying that rotation period
increases with age as expected.
This result is insensitive to the choice of bin position and size.
Black lines show gyrochrones from the \citet{angus2019} gyrochronology model,
which projects the rotation-color relation of Praesepe to longer rotation
periods over time.
These gyrochrones are plotted at 0.5, 1, 1.5, 2, 2.5, 4 and 4.57 Gyr.
At the youngest ages, these gyrochrones describe the data well: the palest
yellow (youngest) stars with the lowest velocity dispersions all fall close to
the 0.5 Gyr gyrochrone.
% Lines of constant age (isochrones) appear to follow the shape of the
% Praesepe-based gyrochronology model (black dotted line) at young ages.
% However, at older ages it appears that the relation between rotation period
% and \teff\ flattens out, until eventually rotation period {\it decreases} with
% decreasing effective temperature at a given age.
However, although the 0.5 Gyr and 1 Gyr gyrochrones also trace constant
velocity dispersion/age among the field stars, by 1.5 Gyr the gyrochrones
start to {\it cross} different velocity dispersion regimes.
For example, the 1.5 Gyr gyrochrone lies on top of stars with velocity
dispersions of around 10-11 kms$^{-1}$ at 5000-5500K and stars with $\sim$15
\kms\ velocity dispersions at 4000-4500K.
The gyrochrones older than 1.5 Gyr also cross a range of velocity dispersions.
If these were true isochrones however, they should follow lines of constant
velocity dispersion.
At ages older than around 1 Gyr, it appears that gyrochrones should have a more
flattened, or even inverted relationship between rotation period and effective
temperature than these Praesepe-based models.
% For late K-dwarfs older than around 1 Gyr, it appears that lower-mass stars
% rotate more rapidly than higher-mass stars.
% This is the opposite of the trend that is observed in young open clusters such
% as Praesepe and the Hyades.

The bottom panel of figure \ref{fig:vplot} shows velocity dispersion as a
function of rotation period and {\it mass}, obtained from the Kepler Input
Catalog \citep{brown2011}, with gyrochrones from the \citep{spada2019}
gyrochronology model shown as white lines.
These gyrochrones are also plotted at 0.5, 1, 1.5, 2, 2.5, 4 and 4.57 Gyr.
Although these models do not trace lines of constant velocity dispersion at
0.5 and 1 Gyr, they do appear to reproduce trends in the data at ages of 1.5
Gyr and older.
The \citet{spada2019} models appear to qualitatively agree with the data:
their rotation period-\teff\ relation flattens out over time and eventually
inverts.

@@

To investigate whether mass-dependent heating could be acting on the \kepler\
sample, we selected late K and M dwarfs observed by both \kepler\ and \gaia,
whose MS lifetimes exceed around 11 Gyr and are therefore representative of
the initial mass function.
We could not perform this analysis on the \mct\ sample, because it only
includes stars with {\it detectable} rotation periods, and since lower-mass
stars stay active for longer it is likely that it contains a strong mass-age
correlation.
We selected all \kepler\ targets with dereddened \gaia\ \gcolor\ colors
greater than 1.2 (corresponding to an effective temperature $\lesssim$
4800 K) and absolute \gaia\ $G$-band magnitudes $<$ 4.
We also eliminated visual binaries by removing stars above a 6th order
polynomial, fit to the MS on the \gaia\ CMD (similar to the one shown in
figure \ref{fig:age_gradient}).
We then applied the quality cuts described above in section
\ref{sec:the_data}.
To search for evidence of mass-dependent heating we calculated the (\vb)
velocity dispersion of stars in effective temperature bins.
Sigma clipping was performed at 3$\sigma$ to remove high and low velocity
outliers before calculating the standard deviation of stars in each bin.
These extreme velocity outliers may be very old late K and M dwarfs, or they
result from using \vb\ instead of \vz, which introduces additional velocity
scatter.

@@

Having found no strong evidence for mass-dependent heating, we next tested
the validity of \vb\ as a proxy for \vz\ in more detail.
At a galactic latitude, $b$, of zero, $v_b=v_z$, however for increasing values
of $b$, this equivalence becomes an approximation that grows noisier with $b$.
To test the validity of the \vb$\sim$\vz\ approximation over a range of
latitudes we downloaded stellar data from the \Gaia\ Universe Model Snapshot
(GUMS) simulation -- a simulated \Gaia\ catalog \citep{robin2012}.
We downloaded stars from four pointings in the \kepler\ field with galactic
latitudes of around 5\degrees, 10\degrees, 15\degrees, and 20\degrees, out to
a limiting magnitude of 16 dex, and calculated their \vz\ and \vb\ velocities.
The relationship between \vz\ and \vb\ is close to 1:1, with \vz\ greater than
\vb\ by around 4.5 kms$^{-1}$ at $b=5$, due to the Sun's own motion in the
Galaxy.
We subtracted this offset and examined the residuals of the \vz\ -- \vb\
relationship to investigate the variance as a function of Galactic latitude
(shown in figure \ref{fig:vb_vz}).
We found that \vb\ is drawn from a heavy-tailed distribution, centered on \vz,
with standard deviation increasing with $b$ (see figure \ref{fig:vb_vz}).
The standard deviation of \vz-\vb\ was around 3kms$^{-1}$ at $b \sim 5^\circ
$, 4kms$^{-1}$ at 10$^\circ$, 6kms$^{-1}$ at 15$^\circ$, and 9kms$^{-1}$ at
20$^\circ$.

@@

Because of the noisy relationship between \vb\ and \vz\, in this paper we do
not attempt to convert velocity dispersion (\sigmavb) into an age via an
age-velocity dispersion relation (AVR) \citep[\eg][]{holmberg2009}.
Although we find that \sigmavb\ can be used to rank groups of stars by age, a
more careful analysis that includes formal modeling of the \vb\ -- \vz\
relationship will be needed to calculate absolute ages.
% For example, the velocity distributions could be modeled as a mixture of
% Gaussians in order to account for the additional velocity dispersion caused by
% the \vz-\vb\ transformation.
% The RVs of most of these stars will become available in \Gaia\ DR3, allowing
% calculations of \vz, which can be used to calculate more reliable ages via an
% AVR and suitable modeling of the selection function.
